---
title: "asbur_chapter4"
author: "Ian Bates"
date: "2/24/2022"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(remedy)
remedy_opts$set(name = "Bates")
```

##### Example 1: Hypothetical Gdf5 gene expression. To illustrate the Ztest we will look at the Gdf5 gene from the Golub et al. (1999) data2. The Gdf5 expression values are contained in row 2058. A quick search through the NCBI site makes it likely that this gene is not directly related to leukemia. Hence, we may hypothesize that the population mean of the ALL expression values equals zero. Accordingly, we test H0 : µ = 0 against Ha : µ 6= 0. For the sake of illustration we shall pretend that the population standard deviation σ is known to be equal to 0.25. The test statistic z-value can be computed as follows:

```{r remedy001}


data(golub, package = "multtest")


gol.fac <- factor(golub.cl,
  levels = 0:1,
  labels = c("ALL", "AML")
)

sigma <- 0.25
n <- 27
mu0 <- 0
x <- golub[2058, gol.fac == "ALL"]
z.value <- sqrt(n) * (mean(x) - mu0) / sigma
z.value

2 * pnorm(-abs(z.value), 0, 1)
```




```{r Bates001, tidy=TRUE}

f <- function(x) {
  dnorm(x, 0, 1)
}
x1 <- seq(-4, -1.96, 0.01)
y1 <- dnorm(x1, 0, 1)
x2 <- seq(-1.96, 1.96, 0.01)
y2 <- dnorm(x2, 0, 1)
x3 <- seq(1.96, 4, 0.01)
y3 <- dnorm(x3, 0, 1)
plot(f,
  -4,
  4,
  cex.lab = 1.5,
  xlab = "x",
  ylab = "Normal probability density function f(x)"
)
polygon(c(-4, x1, -1.96), c(0, y1, 0), col = "red")
polygon(c(-1.96, x2, 1.96), c(0, y2, 0), col = "lightblue")
polygon(c(1.96, x3, 4), c(0, y3, 0), col = "red")
arrows(-3, 0.15, -3, 0.03)
text(-3, 0.23, "Rejection")
text(-3, 0.20, "Region")
text(-3, 0.17, expression(alpha / 2))
arrows(3, 0.15, 3, 0.03)
text(3,0.23,"Rejection")
text(3,0.20,"Region")
text(3,0.17,expression(alpha/2))
text(0,0.23,"Acceptance")
text(0,0.20,"Region")
```




##### Example 2: Confidence interval. Using the data from Example 1, the 95% confidence interval given by Equation 4.1 can be computed as follows:4

```{r}

mean(x)+c(-1,1) * qnorm(c(0.975),0,1)*0.25/sqrt(n)


library(TeachingDemos)
z.test(x,mu=0,sd=0.25)

```



```{r}

ci.examp(
  mean.sim = 0,
  sd = 1,
  n = 25,
  reps = 100,
  method = "z",
  lower.conf = 0.025,
  upper.conf = 0.975
)

```
##### 4.1.2 One-sample t-test In almost all research situations, the population standard deviation σ is unknown - and so the above Z-test is usually not applicable. In such cases, t-tests are very useful for testing H0 : µ = µ0 against Ha : µ 6= µ0, where the sample standard deviation s is used in lieu of the population standard deviation σ. The test is based on the test statistic t-value defined by t = √ n(x−µ0)/s. The corresponding p-value is defined by 2·P(Tn−1 ≤ −|t|). As with the Z-test, H0 is not rejected if the p-value is larger than the significance level and H0 is rejected if the p-value is smaller than the significance level. Equivalently, if t falls in the acceptance region (t0.025,n−1, t0.975,n−1), then H0 is not rejected and otherwise it is. For n = 6, the acceptance and rejection regions are illustrated in Figure 4.2. We can create Figure 4.2 using the dt(), polygon(), arrows(), seq(), mtext(), and text() functions:



```{r}


f <- function(x) {
  dt(x, 5)
}
x1 <- seq(-4, qt(0.025, 5), 0.01)
y1 <- f(x1)
x2 <- seq(qt(0.025, 5), qt(0.975, 5), 0.01)
y2 <- f(x2)
x3 <- seq(qt(0.975, 5), 4, 0.01)
y3 <- f(x3)
plot(f,-4,
     4,
     xlab = "x",
     ylab = "t-Distribution probability density function f(x)")
polygon(c(-4, x1, qt(0.025, 5)), c(0, y1, 0), col = "red")
polygon(c(qt(0.025, 5), x2, qt(0.975, 5)), c(0, y2, 0), col = "lightblue")
polygon(c(qt(0.975, 5), x3, 4), c(0, y3, 0), col = "red")
arrows(-3, 0.15, -3, 0.03)
text(-3, 0.23, "Rejection")
text(-3, 0.20, "Region")
text(-3, 0.17, expression(alpha / 2))
arrows(3, 0.15, 3, 0.03)
text(3, 0.23, "Rejection")
text(3, 0.20, "Region")
text(3, 0.17, expression(alpha / 2))
text(0, 0.23, "Acceptance")
text(0, 0.20, "Region")
mtext(expression(t[0.025]),
      side = 1,
      at = qt(0.025, 5),
      col = "red")
mtext(expression(t[0.975]),
      side = 1,
      at = qt(0.975, 5),
      col = "red")

```

##### Example 1: Actual Gdf5 gene expression. Let’s test H0 : µ = 0 against Ha : µ 6= 0 for the ALL population mean of the Gdf5 gene expressions. The latter are collected in row 2058 of the golub data. The t-value is computed as follows:
