---
title: "Homework 3"
author: "Ian Bates"
date: "2/21/2022"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Binomial distribution. Let X be binomially distributed with n = 60 and p = 0.4. Compute the following.


### (a)
#### $$P(X = 24) $$
```{r}

dbinom(x=24, size = 60, prob = 0.4)
```

#### $$P(X ≤ 24)$$

```{r}
pbinom(q=24, size = 60, prob = 0.4)

```

#### $$P(X ≥ 30)$$

```{r}

pbinom(q=29, size=60, prob=0.4, lower.tail = FALSE)

```
### (b) 
#### $$P(20 ≤ X ≤ 30)$$
```{r}

sum(dbinom(x=20:30, size = 60, prob = 0.4))
```
#### $$P(20 ≤ X)$$
```{r}

pbinom(q=19, size=60, prob=0.4, lower.tail = FALSE)

```


### (c) 
#### $$P(20 ≤ X or X ≥ 40)$$

```{r}

pbinom(q=19, size=60, prob=0.4, lower.tail = FALSE)


```
#### $$P(20 ≤ X and X ≥ 10)$$


```{r}

pbinom(q=9, size=60, prob=0.4, lower.tail = FALSE)

```
### (d) The mean and standard deviation of X.


```{r}
n = 60
p = 0.4

exampleMean <- n*p
exampleMean

variance <- sqrt(n*p*(1-p))
variance

```


### (e) The quantiles:
#### $$x_{0.025}$$

```{r}
qbinom(p = 0.025, size = 60, prob = 0.4)
```
#### $$x_{0.025}$$
```{r}
qbinom(p = 0.5, size = 60, prob = 0.4)
```


#### $$x_{0.975}$$
```{r}
qbinom(p = 0.975, size = 60, prob = 0.4)
```



# 2. Standard normal distribution. Compute the following probabilities and quantiles

### $$(a) P(1.6 < Z < 2.3)$$

```{r}

pnorm(2.3) - pnorm(1.6)

```
### b) $$P(Z < 1.64)$$
```{r}

pnorm(1.64)

```
### c) $$P(−1.64 < Z < −1.02)$$.
```{r}

pnorm(-1.02) - pnorm(-1.64)

```
### (d) $$ P(−1.96 < Z < 1.96) $$
```{r}

pnorm(1.96) - pnorm(-1.96)

```

### (e) $$ P(−1.96 < Z < 1.96) $$
```{r}

pnorm(1.96) - pnorm(-1.96)

```

### (f) The quantiles
#### $$ z_{0.025} $$
```{r}

qnorm(0.025) 

```

#### $$ z_{0.05} $$
```{r}

qnorm(0.05) 

```

#### $$ z_{0.5} $$
```{r}

qnorm(0.5) 

```
#### $$ z_{0.95} $$
```{r}

qnorm(0.95) 

```
#### $$ z_{0.975} $$
```{r}

qnorm(0.975) 

```

# 3. Normal distribution. Compute for X distributed as N (10, 2) the following probabilities and quantiles.

### (a) $$P(X < 12)$$

```{r}

pnorm(12, 10, 2)

```
### (b) $$P(X > 8)$$


```{r}

1 - pnorm(8, 10, 2)

```



### (c) $$P(9 < X < 10, 5)$$

```{r}

pnorm(10.5, 10, 2) - pnorm(9, 10, 2)

```


### (d) The quantiles

#### $$x_{0.025}$$

```{r}

qnorm(0.025, 10, 2)

```


#### $$x_{0.5}$$

```{r}

qnorm(0.5, 10, 2)

```


#### $$x_{0.975}$$

```{r}

qnorm(0.975, 10, 2)

```


# 4. t-distribution. Compute the following probabilities and quantiles for the t6 distribution.


### (a) $$P(t_{6} < 1)$$

```{r}

pt(1, 6)

```



### (b) $$P(t_{6} > 2)$$

```{r}

1- pt(2, 6)

```


### (c) $$ P(−1 < t_{6} < 1) $$

```{r}

pt(1, 6) - pt(-1,6)

```

### (d) $$ P(−2 < t_{6} < −2) $$

```{r}

pt(-2, 6) - pt(-2,6)

```

### (e) The quantiles 
#### $$t_{0.025}$$
```{r}
qt(0.025, 6)
```

#### $$t_{0.5}$$
```{r}
qt(0.5, 6)
```

#### $$t_{0.975}$$
```{r}
qt(0.975, 6)
```

#  5. F distribution. Compute the following probabilities and quantiles for the F8,5 distribution.


### (a) $$ P(F_{8,5} < 3)$$
```{r}
pf(3 , 8,5)

```

### (b) $$ P(F_{8,5} > 4)$$
```{r}
1-pf(4 , 8,5)

```

### (c) $$ P(1 < F_{8,5} < 6) $$
```{r}
pf(6 , 8,5) - pf(1 , 8,5) 

```


### (d) The quantiles
#### $$ f_{0.025} $$
```{r}
qf(0.025 , 8,5) 

```
#### $$ f_{0.5} $$
```{r}
qf(0.5 , 8,5) 

```

#### $$ f_{0.975}$$
```{r}
qf(0.975 , 8,5) 

```

#  6. Chi-squared distribution. Compute the following for the chi-squared distribution with 10 degrees of freedom

### (a) $$P(X_{10}^2 < 3)$$

```{r}

pchisq(3,10)

```
### (b) $$ P(X_{10}^2 > 4) $$

```{r}

pchisq(4, 10, lower.tail = FALSE)

```
### (c) $$ P(1 < X_{10}^2 < 6)$$

```{r}

pchisq(6,10) - pchisq(1,10) 

```

### (d) The quantiles 
#### $$ x_{0.025}$$
```{r}

qchisq(0.025,10) 

```
#### $$x_{0.5}$$ 
```{r}

qchisq(0.5,10) 

```

#### $$x_{0.975}$$
```{r}

qchisq(0.975,10) 

```
# 7. Purines in microRNAs. Suppose that for certain microRNA of size 20 the probability of a purine is binomially distributed with probability 0.7.



### (a) What is the probability of 14 purines?
```{r}
mrnaSize <- 20 
purineProb <- 0.7

dbinom(14, size = mrnaSize, prob = purineProb)


```
### (b) What is the probability of less than or equal to 14 purines

```{r}

pbinom(14, size = mrnaSize, prob = purineProb)
```
### (c) What is the probability of strictly more than 10 purines?

```{r}

pbinom(10, size = mrnaSize, prob = purineProb, lower.tail = FALSE)
```

### (d) What is the probability that there are between 10 and 15 purines, inclusive?


```{r}

pbinom(15, size = mrnaSize, prob = purineProb) - pbinom(9, size = mrnaSize, prob = purineProb)

```

### (e) How many purines do you expect? In other words: What is the mean of the distribution?


```{r}

distributionMean <- mrnaSize*purineProb
distributionMean

```
### (f) What is the standard deviation of the distribution?

```{r}

variance <- sqrt(mrnaSize*purineProb*(1-purineProb))
variance

```

# 8. Zyxin gene expression. The distribution of the expression values of the ALL patients on the Zyxin gene are distributed according to $$N(1.6, 0.4^2)$$


### (a) Compute the probability that the expression values are smaller than 1.2.

```{r}
normMean <- 1.6
normStd <- 0.4

pnorm(1.2,mean = normMean, sd=0.4)

```

### (b) What is the probability that the expression values are between 1.2 and 2.0?

```{r}
pnorm(2, mean = normMean, sd = normStd) - pnorm(1.2, mean = normMean, sd = normStd) 

```

### (c) What is the probability that the expression values are between 0.8 and 2.4?


```{r}
pnorm(2.4, mean = normMean, sd = normStd) - pnorm(0.8, mean = normMean, sd = normStd) 

```


### (d) Compute the exact values for the quantiles 
#### $$x_{0.025}$$ 

```{r}
qnorm(0.025, mean = normMean, sd = normStd)

```
#### $$x_{0.975}$$

```{r}
qnorm(0.975, mean = normMean, sd = normStd)

```


### (e) Use rnorm to draw a sample of size 1000 from the population and compare the sample mean and standard deviation with that of the population.


```{r}

normD <- rnorm(1000, mean = normMean, sd = normStd)
mean(normD)

sd(normD)

```

# 9. Some computations on the Golub et al. (1999) data.


### (a) Take µ = 0 and compute the t-values for the ALL gene expression values. Find the three genes with largest absolute t-values.

```{r}

data(golub, package = "multtest")

golubFactor <- factor(golub.cl,
                      levels = 0:1,
                      labels = c("ALL", "AML"))


ALLPatientCount <- ncol(golub[, golubFactor == "ALL"])

tValues <-
  apply(golub[, golubFactor == "ALL"], 1, function(x)
    sqrt(ALLPatientCount) * (mean(x) - 0) / sd(x))

orderedTValues <- order(tValues, decreasing  = TRUE)

tValues[orderedTValues[1:3]]

golub.gnames[orderedTValues[1:3], 2]

```
### (b) Compute per gene the ratio of the variances for the ALL over the AML patients. How many are between 0.5 and 1.5?
```{r}
data(golub, package = "multtest")
golubFactor <- factor(golub.cl,
                      levels = 0:1,
                      labels = c("ALL", "AML"))

golubALL <- golub[, golubFactor == "ALL"]
golubAML <- golub[, golubFactor == "AML"]

sdALL <- apply(golubALL, 1, sd)
sdAML <- apply(golubAML, 1, sd)

sdAMLsdALLRatio <- sdALL / sdAML

sum(sdAMLsdALLRatio >= 0.5 & sdAMLsdALLRatio <= 1.5, na.rm=TRUE)
```




