---
title: "chapter_8_ASBUR"
author: "Ian Bates"
date: "4/3/2022"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Classification trees with the ALL data. Perform recursive partitioning with the rpart() function. Like Example 5, using the ALL data, we want to predict the diagnosis of B-cell state B1, B2, or B3 from the gene expressions. 
```{r, warning=FALSE, message = FALSE}
# BiocManager::install("hgu95av2.db")
library(ALL)
data(ALL)
library("hgu95av2.db")
library(rpart)
library(rpart.plot)

ALLB123 <- ALL[,ALL$BT %in% c("B1","B2","B3")]

```


### (a) Select the genes with different means over the patient groups with an ANOVA p-value smaller than 0.001.

```{r}

anova.pValue <-
  apply(exprs(ALLB123), 1, function(x)
    anova(lm(x ~ ALLB123$BT))$Pr[1])

names <- featureNames(ALL)[anova.pValue < 0.001]
names
```

### (b) Using rpart(), find a manner to identify the best 2 genes to predict the B-cell state.

```{r}
symb <- mget(names, env = hgu95av2SYMBOL)
ALLBTnames <- ALLB123[names,]

probeData <- as.matrix(exprs(ALLBTnames))
row.names(probeData) <- unlist(symb)
diagnosed <- factor(ALLBTnames$BT)
rpartFit <- rpart(diagnosed ~ ., data = data.frame(t(probeData)))
rpartFit$variable.importance[1:2]


```




### (c) Use rpart() again to construct the best overtall classification tree to predict the B-cell state.

```{r}


prp(
  rpartFit,
  branch.lwd = 4,
  branch.col = "blue",
  extra = 101
)



```


### (d) How do the trees and misclassification rates differ?

```{r}

predicted.class <- predict(rpartFit, type = "class")
predicted.probabilities <- predict(rpartFit, type = "prob")

out <-
  data.frame(predicted.probabilities,
             predicted.class,
             diagnosis = factor(ALLBTnames$BT))
print(head(out, 10), digits = 2)
rpartPredict <- predict(rpartFit, type = "class")
table(rpartPredict, diagnosed)


```
> As we can see in the table, we predicted B1 with 2 false positives and 2 false negatives, 
B2 had 1 false positive and 7 false negatives, and with B3 we had 0 false positives and 6 false negatives. This is likely due to the low P value used in the ANOVA test to reduce the numebr of genes.

