---
title: "Midterm_Bates"
author: "Ian Bates"
date: "3/7/2022"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = "center")
library(reticulate)
use_python('C:/dev/python/python.exe')
```



## Problem 1 - Suppose that trees are distributed in a forest according to a two-dimensional Poisson process with lambda the expected number of trees per acre given by 80.

### (a)(3pts) What is the probability that in a certain quarter acre plot there will be at most 16 trees?
```{r}

lambda.per_acre <- 80
lambda.quarter_acre <- 0.25 * lambda.per_acre
answer.1.a <- ppois(16, lambda = lambda.quarter_acre)

paste("The probability 16 trees in a quarter acre: ",answer.1.a)
```



### (b)(2pts) If the forest covers 85,000 acres what is the expected number of trees in the forest?

```{r}

forest.acres <-85000
answer.1.b <- forest.acres * lambda.per_acre

paste("Expected Trees:", answer.1.b)
```




## Problem 2 Suppose the diameter at breast height (in.) of a certain type of tree is distributed normally with a mean of 8.8 and a standard deviation of 2.8 as suggested in Aedo-Ortiz, D. M., Olsen, E. D., & Kellogg, L. D. (1997). Simulating a harvester-forwarder softwood thinning: a software evaluation. Forest Products Journal, 47(5), 36.

### (a)(2pts) What is the probability that the diameter of a randomly selected tree will be at least 10 inches? Will exceed 10 inches.

```{r}
p2.sd <- 2.8
p2.mean <- 8.8
p2.diameter <- 10

x <- seq(0,20, 0.1)

dnormFun <- function(x) { dnorm(x,p2.mean,p2.sd) }
plot(dnormFun,
     # function
     0,
     # start
     20,
     # end
     xlab = "x",
     ylab = "Normal probability density function f(x)",
     col = "red")

p2.greater_than_10 <- 1 - pnorm(10 , mean = p2.mean, sd = p2.sd)

paste("Probability of tree diameter exceeding 10: ", p2.greater_than_10)


```


### (b)(3pts) What is the value of c so that the interval (8.8-c, 8.8+c) contains 98% of all diameter values

>Since we want 98% of the values to be in between $8.8-c$ and $8.8+c$, I'm going to look for that value that is at the 99% percentile for the upper bound


```{r}

p2.percentile <- 0.99

p2.val <- qnorm(p2.percentile, mean = p2.mean, sd = p2.sd)

p2.c <- p2.val - p2.mean

p2.c

# validation....

pnorm(p2.mean + p2.c, mean = p2.mean, sd = p2.sd) - pnorm(p2.mean - p2.c, mean =
                                                            p2.mean, sd = p2.sd)


```


## Problem 3 - Suppose that a group of 100 men aged 60-64 received a new flu vaccine in 1986 and that 5 of them died within the next year. Assume that a 1986 mortality table tells us that men aged 60-64 have an approximate probability of death in the next year of .020.


### (a)(3pts) How unusal is it for the 5 men to have died? [Hint: Be careful to review my discussions on assessing extreme events. Also, use the binomial model.]

```{r}

# DELETE THIS!!! pg. 96: https://www.usu.edu/math/corcoran/classes/old/11spring5100/readings/rosner04.pdf
p3.n <- 100
p3.event <- 5
p3.p <- 0.020

```

```{r, include=FALSE}
par(mfrow = c(1, 1))

x <- 0:12
plot(
  x,
  dbinom(x, size = p3.n, prob = p3.p),
  type = "h",
  col = "blue",
  lwd = 4,
  xlab = "x",
  ylab = "Binomial probability mass function f(x)"
)

binomialCDF = stepfun(x, c(0, pbinom(x, size = p3.n, prob = p3.p)))

plot(
  binomialCDF,
  col = "red",
  vertical = FALSE,
  lwd = 4,
  xlab = "x",
  ylab = "Binomial cumulative distribution function F(x)",
  main = NULL
  )

```

>When we're assessing the anomalous significance of `r p3.event` people dying in a population, we are going to assess the right tailed test $Pr(X>=`r p3.event`|X~binom(`r p3.n`, `r p3.p`))$

>And because $pbinom(x, ... , lower.tail = FALSE)$ computes percentages for values strictly greater than x, we will have to subtract 1 from our event number `r p3.event`.

```{r}
pbinom(p3.event - 1, size = p3.n, prob = p3.p, lower.tail = FALSE)

```
> All semester, we have been utilizing p < 0.05 as an indicator for statistical significance, I would say that a p value of `r pbinom(p3.event - 1, size = p3.n, prob = p3.p, lower.tail = FALSE)` not sufficient to warent any concern.

### (b)(2pts) Suppose 10 men had died. Should we be concerned?

```{r}

p3.new_event <- 10

pbinom(p3.new_event-1, size = p3.n, prob = p3.p, lower.tail = FALSE)


```

>Now for this instance where 10 people die, significance of $Pr(X>=`r p3.new_event`|X~binom(`r p3.n`, `r p3.p`))$ gives us a tremendously small percentage $`r pbinom(p3.new_event-1, size = p3.n, prob = p3.p, lower.tail = FALSE)`$ which is a cause for concern.


## Problem 4 The incidence of a certain type of chromosome defect in the U.S. adult male population is beleived to be 1 in 75. A random sample of 800 individuals in U.S. penal institutions reveals 16 who have such defects. Can it be concluded that the incidence rate of this defect among prisoners differs from the preseumed rate of the entire adult population.

### (a)(3pts) State and test the relevant hypothesis using α=.05. What type of error might you have made in reaching a conclusion?


>The null hypothesis is $$H_{0} = 1/75$$

```{r}
# chapter 8, problem 67
# less useful
# https://www.chegg.com/homework-help/incidence-certain-type-chromosome-defect-us-adult-male-popul-chapter-8-problem-75e-solution-9780538733526-exc
# https://www.chegg.com/homework-help/questions-and-answers/incidence-certain-type-chromosome-defect-us-adult-male-population-believed-1-75-random-sam-q14864927
p4.rate <- 1/75
p4.sample <- 800
p4.incident <- 16
p4.alpha <- 0.5


binom.test(x=p4.incident, n = p4.sample, p=p4.rate, alternative = 'two.sided')

p4.binomTest <- binom.test(x=p4.incident, n = p4.sample, p=p4.rate, alternative = 'two.sided')

p4.binomTest$p.value
```

>With a p value of `r p4.binomTest$p.value`, we fail to reject the null hypothesis.  This error is a type 2 error.


### (b)(2pts) What P-value is associated with this test. Based on this P-value could H_0 be rejected at a significance level of .2

> Yes, we would reject the null hypothesis if we we have a significance value of 0.2 as `r p4.binomTest$p.val` < 0.2



## Problem 5 Two different fish attractors were compared during 16 time periods spanning 4 years. (Wilbur, R. L. (1978). Two types of fish attractors compared in Lake Tohopekaliga, Florida. Transactions of the American Fisheries Society, 107(5), 689-695.) I present the data as a set of ordered pairs where in each case the first entry is the “pipe attractor” and the second entry is the “brush attractor”
{(6.64,9.73), (7.89,8.21),(1.83,2.17),(.42,.75),(.85,1.61),(.29,.75),(.57,.83),(.63,.56),(.32,.76),(.37,.32),(.00,.48),(.11.,.52),(4.86,5.38),(1.80,2.33),(.23,.91,),(.58,.79)}

### (a)(4 pts) Perform the appropriate paired parameteric t-based test to compare the means.

```{python, include=FALSE}
base_str = '(6.64,9.73),(7.89,8.21),(1.83,2.17),(.42,.75),(.85,1.61),(.29,.75),(.57,.83),(.63,.56),(.32,.76),(.37,.32),(.00,.48),(.11,.52),(4.86,5.38),(1.80,2.33),(.23,.91),(.58,.79)'

split_str = base_str.split('),')
split_str = [x.replace('(','') for x in split_str]
split_str = [x.replace(')','') for x in split_str]
split_str = [x.split(',') for x in split_str]
pipe_attr = []
brush_attr = []
pipe_r = ''
brush_r = ''
for val in split_str:
    pipe_attr.append(float(val[0]))
    brush_attr.append(float(val[1]))

pipe_r = 'c('+ ','.join([f"{x:.2f}" for x in pipe_attr]) + ')'
brush_r = 'c('+ ','.join([f"{x:.2f}" for x in brush_attr]) + ')'
print(pipe_r)
print(brush_r)

```


```{r}
# https://www.chegg.com/homework-help/questions-and-answers/5-two-different-fish-attractors-compared-16-time-periods-spanning-4-years-wilbur-r-l-1978--q45849673
# https://www.chegg.com/homework-help/questions-and-answers/1-10-two-types-fish-attractors-one-made-vitrified-clay-pipes-cement-blocks-brush-used-16-d-q38083599
pipe  <- c(6.64,7.89,1.83,0.42,0.85,0.29,0.57,0.63,0.32,0.37,0.00,0.11,4.86,1.80,0.23,0.58)
brush <- c(9.73,8.21,2.17,0.75,1.61,0.75,0.83,0.56,0.76,0.32,0.48,0.52,5.38,2.33,0.91,0.79)

p5.t<-t.test(pipe, brush, paired=TRUE)
p5.t

```

> If we look at the p value for the paired T-Test, we will see that `r p5.t$p.val`
< 0.05, which means that we can reject the null hypothesis of equal means




### (b)(6 pts) What happens if the two independent samples t-test is used? Make sure to perform all appropriate tests.


```{r}

p5.var <- var.test(pipe, brush)
p5.var

```

>Since the F test for equal variances shows us that we cannon reject the null hypothesis of equal variances, becuase `r p5.var$p.value` > 0.05, we will utilize the t.test with var=TRUE.


```{r}

t.test(pipe, brush, var.equal = TRUE)


```


## Problem 6 The following article (Thomas, H. V., & Simmons, E. L. L. A. M. A. E. (1969). Histamine content in sputum from allergic and nonallergic individuals. Journal of applied physiology, 26(6), 793-797) reports the following data on sputum histamine levels (μg/g dry weight of sputum) for a sample of 9 individuals classified as allergics and 13 individuals classified as nonallergics.

Allergics = {67.7, 39.6,1651.0, 100.0, 65.9,1112.0,31.0, 102.4,64.7}
Nonallergics = {34.3, 27.3, 35.4, 48.1, 5.2, 29.1, 4.7, 41.7, 48.0, 6.6, 18.9, 32.4, 45.5}

### (a)(4 pts) Ascertain the normality of these datasets using the Shapiro Wilks test.
```{r}
p6.allergics = c(67.7, 39.6,1651.0, 100.0, 65.9,1112.0,31.0, 102.4,64.7)
p6.nonallergics = c(34.3,27.3,35.4,48.1,5.2,29.1,4.7,41.7,48.0,6.6,18.9,32.4,45.5)


shapiro.test(p6.allergics)
shapiro.test(p6.nonallergics)


```

### (b)(6pts) You should have seen that one of these failed the test. Perform the appropriate nonparametric test to compare the means. Does the P-value suggest that we can reject at the .01 level?
```{r}
# https://www.chegg.com/homework-help/questions-and-answers/1-data-table-43-subset-data-obtained-thomas-simmons-1969-investigated-relation-sputum-hist-q45212501

p6.w <- wilcox.test(p6.allergics, p6.nonallergics)
p6.w
```

> With a p value of `r p6.w$p.val`, we would reject the null hypothesis


## Problem 7 (10 pts) In a well-known genetics article (Yule, G. U. (1923). The progeny, in generations F 12 to F 17 of a cross between a yellow-wrinkled and a green-round seeded pea; a report on data afforded by experiments initiated by the late AD Darbishire, MA, in 1905, and conducted by him until his death in 1915. Journal of Genetics, 13(3), 255-331) the early statistician G. U. Yule analyzed data from crossing garden peas. The dominant alleles were Y = yellow color and R = round shape resulting in a double dominant YR. Yule examined 269 4 seed pods resulting from a dihybrid cross and counted the number of YR seeds in each pod. Letting X represent the number of seeds in a randomly selected pod with possible X values of 0, 1, 2, 3, and 4 which are identified with cells 1, 2, 3, 4, 5 of a 1 x 5 table. So a pod with X = 4 would have its count values in the 5th entry of the table.The hypothesis that the Mendelian laws are operative and that genetypes of individual seeds within a pod are independent of one another implies that X has a binomial distribution with n=4 and p=9/16. The null hypothesis is that the Medelian laws hold true and the alternative hypothesis is that they do not. The entry in the table (the observed seed counts) reading from the first to the fifth are {16,45,100,82,26}.The probabilities under Ho are given by the values of the binomial distribution with n=4 and p=9/16 as the number of seeds i obtains values in the set {0,1,2,3,4}. Notice that the observed seed count for the 0th case is 16, the 1 case 45, etc. Use the R function chisq.test to test H0. [Hint - You will have to setup a vector of the observed values and a vector of the expected probabilities. Use the values as detailed above.] Do we reject the null hypothesis at the α level of 0.1?

```{r}
p7.n <- 4
p7.p <- 9 / 16
p7.size <- 269


p7.observed <- c(16, 45, 100, 82, 26)
p7.x <- c(0,1,2,3,4)
p7.expected.freq <- sapply(p7.x, function(x) dbinom(x,size = p7.n, prob = p7.p ))

p7.chi_result <- chisq.test(p7.observed, p=p7.expected.freq)
p7.chi_result

```

>No, we would not reject the null hypothesis as `r p7.chi_result$p.val`>0.1


## Problem 8 A Number of studies have shown lichens (certain plants composed of an algae and a fungus) to be excellant bioindicators of air pollution. The article Bruteig, I. E. (1993). The epiphytic lichen Hypogymnia physodes as a biomonitor of atmospheric nitrogen and sulphur deposition in Norway. Environmental monitoring and assessment, 26(1), 27-47. gives the following data (read from a graph) on x=NO; wt deposition (g N/m^2) and y = lichen N (%dry weight)
x = {0.05,0.10,0.11,0.12,0.31,0.37,0.42,0.58,0.68,0.68,0.73,0.85,0.92} 
y = {0.48,0.55,0.48,0.50,0.58,0.52,1.02,0.86,0.86,1.00,0.88,1.04,1.70}

### (a)(4pt) Fit a linear model to the data. What is the value of the slope and y intercept?
```{r}
p8.NO <- c(0.05,0.10,0.11,0.12,0.31,0.37,0.42,0.58,0.68,0.68,0.73,0.85,0.92)
p8.N <- c(0.48,0.55,0.48,0.50,0.58,0.52,1.02,0.86,0.86,1.00,0.88,1.04,1.70)
p8.df <- data.frame(NO=p8.NO, N=p8.N)

p8.ls <- lm(p8.N~p8.NO)
p8.ls <- lm(N~NO, data=p8.df)

cat('The intercept is ', p8.ls$coefficients[1], ' and the slope is ', p8.ls$coefficients[2])


summary(p8.ls)

plot(
  p8.NO,
  p8.N,
  pch = 19,
  cex.lab = 1.5,
  col = "blue",
  xlab = "NO; wt deposition (g N/m^2)",
  ylab = "lichen N (%dry weight)", 
  ylim= c(0.25,2)
)


predicted <- predict(p8.ls)


for (i in 1:length(predicted)) {
  segments(p8.NO[i],
           predicted[i],
           p8.NO[i],
           p8.N[i],
           lwd = 2,
           lty = 2,
           col = "green")
}
abline(p8.ls$coef,
       lwd = 3,
       lty = 2,
       col = "red")


```


### (b)(2pts) Predict lichens N for a NO value of .5 .
```{r}

predict(p8.ls, newdata = data.frame(NO=c(0.5)))

```


### (c)(2pts) Does the model seem to be statistically significant based on the F statistic?

> Yes. Based on the p value from the F statistic 0.0002581 < 0.05

### (d)(2pts) What percentage of the variation of y is explained by the variation in x? [Hint - Do not use the adjusted R-squared.]

> About 0.7175 of the variance is explained by the regression.

## Problem 9 An article Annest, J. L., Pirkle, J. L., Makuc, D., Neese, J. W., Bayse, D. D., & Kovar, M. G. (1983). Chronological trend in blood lead levels between 1976 and 1980.New England Journal of Medicine, 308(23), 1373-1377 gives the following data on y = average blood lead level of white children age 6 months to 5 years and x = amount of lead used in gasoline production (in 1000 tons) for ten 6-month periods.

x = {48,59,79,80,95,95,97,102,102,107} 
y={9.3,11.0,12.8,14.1,13.6,13.8,14.6,14.6,16.0,18.2}.

### (a)(3 pts) Construct normal probability plots for x and y. Might it be reasonable to assume that they come from a bivariate normal distribution?

```{r}

# https://www.chegg.com/homework-help/article-chronological-trend-blood-lead-levels-n-engl-j-med-1-chapter-12-problem-65e-solution-9780534399337-exc

p9.x <- c(48, 59, 79, 80, 95, 95, 97, 102, 102, 107)
p9.y <- c(9.3, 11.0, 12.8, 14.1, 13.6, 13.8, 14.6, 14.6, 16.0, 18.2)

par(mfrow=c(1,2))

qqnorm(
  p9.x,
  pch = 19,
  # cex.lab = 1.5,
  col = "red",
  ylab = "lead used in gasoline sample quantile",
  main = NULL, 
  ylim = c(25,125)
)
qqline(p9.x, col = "blue")

  

qqnorm(
  p9.y,
  pch = 19,
  # cex.lab = 1.5,
  col = "red",
  ylab = "average blood lead level sample quantile",
  main = NULL, 
)
qqline(p9.y, col = "blue")

  
  
```

> Yes, it is reasonable that the they are normally distributed.  There are a few outliers at the end of the distributions, but for the most part, yes the do follow the normal distribution.


### (b)(3 pts) Examine a scatterplot of y versus x. Does it look like there may be some correlation structure?
```{r}
par(mfrow=c(1,1))
p9.ls <- lm(p9.y~p9.x)


plot(p9.x, p9.y, # values
 pch=19, # plot solid circles
 # cex.lab=1.5, # make axis labels big
 xlab="lead used in gasoline ",
 ylab="average blood lead level",
 col="blue")

abline(p9.ls$coef, lwd=3, lty=2, col="red") 


```


> The plot, along with the regression line, do show that there is a linear correlation.

### (c)(4 pts) Does the data provide statistically significant information that there is a linear relationship between blood lead level and the amount of lead used in gasoline production? [Hint: Use an α level of .05 and the R function cor.test]

```{r}

shapiro.test(p9.x)
shapiro.test(p9.y)

p9.cor <- cor.test(p9.x, p9.y)

p9.cor
```

> With the p value `r p9.cor$p.val`< 0.5, we can say that there is a linear relationship between blood lead level and the amount of lead used in gasoline production.


## Problem 10 Suppose that systolic blood pressure, birth weight (oz) and age (days) are measure for a number of infants. Suppose that the data values ae given by 
bp = {89,90,83,77,92,98,82,85,96,95,80,79,86,97,92,88} 
bw = {135,120,100,105,130,125,125,105,120,90,120,95,120,150,160,125} 
age = {3,4,3,2,4,5,2,3,5,4,2,3,3,4,3,3}

### (a)(6pts) Fit a multiple linear regression model given by bp = slope + (m1)(bw) + (m2)(age) + e where e is the error of the model distributed normally with a mean of 0 and a variance of σ^2

```{r}
# https://www.chegg.com/homework-help/questions-and-answers/suppose-systolic-blood-pressure-birth-weight-oz-age-days-measure-number-infants-suppose-da-q45850529
p10.bp <- c(89, 90, 83, 77, 92, 98, 82, 85, 96, 95, 80, 79, 86, 97, 92, 88)
p10.bw <- c(135,120,100,105,130,125,125,105,120,90,120,95,120,150,160,125)
p10.age <- c(3, 4, 3, 2, 4, 5, 2, 3, 5, 4, 2, 3, 3, 4, 3, 3)
p10.df <- data.frame(bp=p10.bp, 
                     bw=p10.bw,
                     age=p10.age)

p10.ls <- lm(bp ~ bw + age, data=p10.df)
summary(p10.ls)


```

>The multiple co


### (b)(2 pts) Predict the bp for a baby with a weight of 8 pounds at 3 days of life.

```{r}

predict(p10.ls, newdata = data.frame(bw=c(8), age=c(3)))

```


## (c)(2 pts) Interpret the regression coefficients of the model. [Hint - See the discussions in ISLR about the linear model of the advertising data.]

## Problem 11 Consider the famous Fisher Iris data.
```{r}
library(datasets)
data(iris)

```


### (a)(5pts) Make a pairs plot of this data coloring the observations based on the species designation using a red, green, blue color scheme and a plot symbol of pch = 21.[Hint: You will need to adjust the figure margin using oma=c(3,3,15) when you call pairs and also call par(xpd=TRUE) prior to calling legend. Position your legend in the bottomright of the plot]

```{r}

oma=c(3,3,15)
pairs(iris, 
      col = c("red", "green", "blue")[unclass(iris$Species)],
      pch = 21,
      oma=c(3,3,3,15)
      )
par(xpd=TRUE)

legend("bottomright", fill = unique(iris$Species), legend = c( levels(iris$Species)))



```




### (b)(5pts)Repeat the exercise of part (a) except produce a parallel coordinates plot.[Hint - It may be impossible to place the legend so it does not obscure the plot a little.]


```{r}

library(MASS)
parcoord(iris[, 1:4],
         col = c("red", "green", "blue")[unclass(iris$Species)],
         oma = c(3, 3, 3, 15))
par(xpd = TRUE)
legend("bottomright",
       fill = unique(iris$Species),
       legend = c(levels(iris$Species)))

```


## Problem 12 Let’s return to the Fisher Iris data.

### (a)(5pts) Use the summary function to explore this data.
```{r}
library(datasets)
data(iris)
summary(iris)

```


### (b)(5pts) Create a 2 x 2 matrix of box plots of the Fisher iris data with the sepal width and sepal length in the first row and the petal width and petal length in the second row. Make sure that the boxplots are conditioned by species.
```{r}

par(mfrow=c(2,2))

boxplot(iris$Sepal.Width ~ iris$Species, # values
           main= "Iris Sepal Width",
           xlab="Species",
           ylab="Sepal Width",
           col=c("purple","green", "blue")
)

boxplot(iris$Sepal.Length ~ iris$Species, # values
           main= "Iris Sepal Length",
           xlab="Species",
           ylab="Sepal Length",
           col=c("purple","green", "blue")
)

boxplot(iris$Petal.Width ~ iris$Species, # values
           main= "Iris Petal Width",
           xlab="Species",
           ylab="Petal Width",
           col=c("purple","green", "blue")
)

boxplot(iris$Petal.Length ~ iris$Species, # values
           main= "Iris Petal Length",
           xlab="Species",
           ylab="Petal Length",
           col=c("purple","green", "blue")
)


```






