---
title: "ProstateCancer"
author: "Ian Bates"
date: "3/9/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
library(cBioPortalData)
library(AnVIL)

```

```{r ,label='investigation'}
library(knitr)
browseVignettes("MultiAssayExperiment")

```

```{r, label='MultiAssayExperiment-doc'}

library(MultiAssayExperiment)
library(GenomicRanges)
library(SummarizedExperiment)
library(RaggedExperiment)
library(jsonlite)
library(survival)
library(survminer)


```
```{r}
# https://waldronlab.io/MultiAssayWorkshop/articles/Ramos_MultiAssayExperiment.html
# https://bioconductor.org/packages/devel/bioc/vignettes/MultiAssayExperiment/inst/doc/QuickStartMultiAssay.htmlA

library(curatedTCGAData)
library(TCGAutils)
library(SingleCellMultiModal)
library(UpSetR)
library(GenomicDataCommons)
```

workshop follow along
https://cbioportal.github.io/2020-cbioportal-r-workshop/

https://cbioportal.github.io/2020-cbioportal-r-workshop/Webinar_Example1.pdf


https://cbioportal.github.io/2020-cbioportal-r-workshop/Example2.pdf

sudy json:
{
    "name": "Prostate Adenocarcinoma (TCGA, PanCancer Atlas)",
    "description": "Prostate Adenocarcinoma TCGA PanCancer data. The original data is <a href=\"https://gdc.cancer.gov/about-data/publications/pancanatlas\">here</a>. The publications are <a href=\"https://www.cell.com/pb-assets/consortium/pancanceratlas/pancani3/index.html\">here</a>.",
    "publicStudy": true,
    "pmid": "29625048,29596782,29622463,29617662,29625055,29625050,29617662,30643250,32214244,29625049,29850653",
    "citation": "TCGA, Cell 2018",
    "groups": "PUBLIC;PANCAN",
    "status": 0,
    "importDate": "2022-01-20 18:25:32",
    "allSampleCount": 494,
    "readPermission": true,
    "studyId": "prad_tcga_pan_can_atlas_2018",
    "cancerTypeId": "prad",
    "referenceGenome": "hg19"
}
```{r}
library(rjson)
study.json <- fromJSON(file='study_info.json')
study.json$studyId




```

```{r, label='intialize-bioportal'}
cbio <- cBioPortal()

cbio


cBioCache()

setCache(
    directory = tools::R_user_dir("cBioPortalData", "cache"),
    verbose = TRUE,
    ask = interactive()
)


study.all_data <- cBioDataPack(
    study.json$studyId,
    names.field = c("Hugo_Symbol", "Entrez_Gene_Id", "Gene"),
    cleanup = TRUE,
    ask = TRUE
)
study.all_data
class(study.all_data)
metadata(study.all_data)
```


```{r, label='cBioPortal-playground'}

show(study.all_data)
length(study.all_data)
names(study.all_data)
ct <- dimnames(study.all_data)

colData(study.all_data)
sampleMap(study.all_data)
metadata(study.all_data)
summary(complete.cases(study.all_data))
colnames(study.all_data)
experiments(study.all_data)

subset_experiment <- study.all_data[, , c('cna', 'log2_cna')]
class(subset_experiment)
experiments(subset_experiment)


?upsetSamples
upsetSamples(study.all_data )
upsetSamples(subset_experiment)


```


```{r, label="extract-assay-sample"}
sample_map_df <- sampleMap(study.all_data["cna"])
t <- sample_map_df$colname
primary_id <- sample_map_df@listData[["primary"]]
length(t)
length(primary_id)
u_p <
u_t <- unique(t)
length(u_t )

class(sample_map_df)


head(sample_map_df)


experiments(study.all_data)

```



```{r, label='combine-assays'}
# getWithColData:  https://rdrr.io/bioc/MultiAssayExperiment/f/vignettes/MultiAssayExperiment.Rmd 
(cna_data <- getWithColData(study.all_data,'cna' ))
cna_data
class(cna_data)


cna_log2_data <- getWithColData(study.all_data, 'log2_cna')
cna_log2_data


combined <- rbind(cna_data, cna_log2_data)
combined

```

```{r, label='subsetting-summary-assays'}

```



```{r}


all_clinical_data = clinicalData(cbio, studyId = study.json$studyId)
head(all_clinical_data)
table(all_clinical_data$TUMOR_TYPE)
table(all_clinical_data$NEW_TUMOR_EVENT_AFTER_INITIAL_TREATMENT)




test_assay <- assays(study.all_data, withDimnames=TRUE)
test_assay2 <- study.all_data@sampleMap@listData[["assay"]]
cna_log2 <- test_assay@listData[["log2_cna"]]
cna <- test_assay@listData[["cna"]]

coldat <- as.data.frame(colData(study.all_data))
table(study.all_data$AGE)
table(study.all_data$ICD_O_3_HISTOLOGY)
table(study.all_data$TISSUE_SOURCE_SITE)
table(study.all_data$PATH_N_STAGE)
table(study.all_data$CANCER_TYPE)
table(study.all_data$CANCER_TYPE_DETAILED)

assays(study.all_data)
all_assays <- assays(study.all_data)

cna_data <- getWithColData(study.all_data,'cna' )
cna_data
rowRanges(cna_data)
colData(cna_data)

cna_log2_data <- getWithColData(study.all_data, 'log2_cna')
cna_log2_data
combined <- rbind(cna_data, cna_log2_data)
combined
browseVignettes("SummarizedExperiment")



?getWithColData
# https://www.bioconductor.org/help/course-materials/2019/BSS2019/04_Practical_CoreApproachesInBioconductor.html
colData(cna_data)

cna_test_data <- as.data.frame(cna_data)
sampleTables(study.all_data)
?sampleTables


# coldat$y <- Surv(study.all_data$, study.all_data$vital_status)

class(cna)
cna['SLC7A11',2]
cna_log2['SLC7A11',2]
cna['UBE2Q2P2',1]
cna_log2['UBE2Q2P2',1]

```
```{r, label='compare-cbioportal-vs-published-data'}
published_data <- read.csv('C:/Users/ianch/GeorgeMason/binf702/r_code/binf702/finalProject/data_curated_trans.csv')
p_data <- data.frame(published_data[,-1], row.names = published_data[,1])


cna_t <- as.data.frame(t(as.matrix(cna)))
c_names <- colnames(cna_t)
c_names <- paste(c_names, "_IB", sep = "")
c_rows <-rownames(cna_t)
c_rows[1]
c_rows <- substr(c_rows, 1, nchar(c_rows)-3)
colnames(cna_t) <- c_names
rownames(cna_t) <- c_rows

total_data <- merge(study.cur_data, cna_log2_t, by.x=0, by.y=0, all=FALSE)
ordered_df <- total_data[,order(colnames(total_data))]

```



```{r}

study.id <- 'prad_tcga_pan_can_atlas_2018'
study.name <- 'Prostate Adenocarcinoma (TCGA, PanCancer Atlas)'
# getStudies(cbio)

study.data <- clinicalData(cbio, studyId = study.id)
study.samples <- allSamples(cbio, studyId = study.id)
study.genes <- getDataByGenes(cbio, studyId = study.id)

molecularProfiles(cbio, studyId = study.id)
sampleLists(cbio, studyId = study.id)
t<-samplesInSampleLists(cbio, sampleListIds = c('prad_tcga_pan_can_atlas_2018_all'))
?getDataByGenes
genePanels(api=cbio)

all_data <- cBioPortalData(cbio, studyId = study.id, genePan)

study.cur_data <- read.csv('C:/Users/ianch/GeorgeMason/binf702/r_code/binf702/finalProject/data_curated_trans.csv')
head(study.cur_data)
ncol(study.cur_data)

study.cur_data[,'ACTRT2']
study.cur_data[,'M_SLC7A11']
study.cur_data[1, 'SLC7A11']
study.cur_data[1,'UBE2Q2P2']
study.cur_data[1,'M_UBE2Q2P2']

cna_log2_t <- as.data.frame(t(as.matrix(cna)))
c_names <- colnames(cna_log2_t)
c_names <- paste(c_names, "_IB", sep = "")
c_rows <-rownames(cna_log2_t)
c_rows[1]
c_rows <- substr(c_rows, 1, nchar(c_rows)-3)
colnames(cna_log2_t) <- c_names
rownames(cna_log2_t) <- c_rows

total_data <- merge(study.cur_data, cna_log2_t, by.x=0, by.y=0, all=FALSE)
ordered_df <- total_data[,order(colnames(total_data))]




study.all_data <- cBioDataPack(
study.id,
names.field = c("Hugo_Symbol", "Entrez_Gene_Id", "Gene"),
cleanup = TRUE,
ask = TRUE
)
```

